An evaluation framework for trajectory inference methods

### 2017-11-24 analysis/paper/paper.Rmd
update paper

```{r, fig.width = 15, fig.height = 10}
n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", exp_id="method_characteristics"))
trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", exp_id="method_characteristics"))

@@

Given[d] this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem. Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art. In this study, we therefore for the first time developed a comprehensive evaluation framework for trajectory inference methods. We test both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge. We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets. To make our framework reusable and extendable, we make use of continuous analysis [@beaulieu-jones_reproducibility_2017], which allows the developers of new methods to easily test their methods and compare them against the state-of-the-art.
# Results
## Evaluation workflow
### Method characterization
```{r}
# Load method characteristics
method_df <- readRDS(derived_file("method_df.rds", exp_id="method_characteristics"))
```

[e]
We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, primarily based on their free availability and the presence of a programming interface. We characterized all methods in four different ways: underlying algorithm(s), implementation quality, prior information and possible trajectory structures (`r ref("fig", "method_characteristics")`). The underlying algorithms can be divided into several subcomponents, and these components are frequently shared between different algorithms. Some components, such as a minimal spanning tree or principal curves for cellular ordering, or PCA or diffusion map for dimensionality reduction are shared by more than 5 methods sFIG?. While not directly related to the quality of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it easier for other developers to adapt the method and extend it further. We therefore looked at the implementation of each method, and assessed their implementation quality using a transparent scoring scheme sNOTE. We first contacted all the authors, to allow them to improve their method before publishing our evaluation. While most methods lack at several metrics (`r ref("fig", "method_characteristics", "a")`), code quality tends to increase over time sFIG. However, some implementation details are lacking for the majority of the methods and should receive extra attention from developers: code availability, documentation and unit testing (`r ref("fig", "method_characteristics", "a")`).

```{r, fig.width = 12, fig.height = 9}
method_characteristics <- readRDS(figure_file("method_characteristics.rds", exp_id="method_characteristics"))

@@

### Data sources

### Score metrics and parameter optimisation

@@

[a]General info about single-cell & trajectories
[b]History of TI, different methods
[c]Common framework of TI, different components, naturally progressing towards the need for an evaluation
[d]Context and general overview of our evaluation
[e]I would move some stuff from this paragraph to supplementary if possible

>>>>

```{r, fig.width = 15, fig.height = 10}
n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", experiment_id="method_characteristics"))
trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", experiment_id="method_characteristics"))

@@

Given[d] this plethora of available trajectory inference methods, it is important that methods are compared, so that users can use the most optimal method available for their problem. Moreover, new methods need to be rigorously tested, so that development can focus on improving the current state-of-the-art. In this study, we therefore for the first time developed a comprehensive evaluation framework for trajectory inference methods. We test both on synthetic data, for which the gold standard is known, and real data, for which in some cases the gold standard can be extracted from expert knowledge. We include a rigorous parameter optimisation, making sure that parameters are optimized while avoiding overfitting on characteristics of specific datasets. To make our framework reusable and extendable, we make use of continuous analysis [@beaulieu-jones_reproducibility_2017], which allows the developers of new methods to easily test their methods and compare them against the state-of-the-art.
# Results
## Evaluation workflow
### Method characterization
```{r}
# Load method characteristics
method_df <- readRDS(derived_file("method_df.rds", experiment_id="method_characteristics"))
```

We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(!is.na(method_df$qc_score))` for evaluation, primarily based on their free availability and the presence of a programming interface. We characterized all methods in four different ways: underlying algorithm(s), implementation quality, prior information and possible trajectory structures (`r ref("fig", "method_characteristics")`). The underlying algorithms can be divided into several subcomponents, and these components are frequently shared between different algorithms. Some components, such as a minimal spanning tree or principal curves for cellular ordering, or PCA or diffusion map for dimensionality reduction are shared by more than 5 methods sFIG?. While not directly related to the quality of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further. We therefore looked at the implementation of each method, and assessed their implementation quality using a transparent scoring scheme sNOTE. After initial scoring we contacted all the authors, to allow them to improve their method before publishing. While most methods lack at several metrics (`r ref("fig", "method_characteristics", "a")`), recent methods tend to have higher code quality sFIG. However, some implementation details are lacking for the majority of the methods and should receive extra attention from developers: code availability, documentation and unit testing (`r ref("fig", "method_characteristics", "a")`).

```{r, fig.width = 12, fig.height = 9}
method_characteristics <- readRDS(figure_file("method_characteristics.rds", experiment_id="method_characteristics"))

@@

### Data sources
![evaluation_overview](../../analysis/figures/evaluation_overview.png)
`r ref("fig", "evaluation_overview", anchor=TRUE)` Overview of the evaluation

We use three types of datasets for our evaluation
### Score metrics and parameter optimisation

@@

[a]General info about single-cell & trajectories
[b]History of TI, different methods
[c]Common framework of TI, different components, naturally progressing towards the need for an evaluation
[d]Context and general overview of our evaluation
